### 2018-06-28 paper/2-background.tex
Work on S2

Build systems automate the execution of simple repeatable tasks for individual
users, as well as for large organisations. In this section we explore the design
space of build systems, using four concrete examples:
\Make~\cite{feldman1979make}, \Shake~\cite{mitchell2012shake},
\Bazel~\cite{bazel}, and \Excel~\cite{advanced_excel}\footnote{\Excel appears
very different to the others but, seen through the lens of this paper, it is
very close indeed.}.
We have carefully chosen these four to illustrate the various axes on which build systems
differ; we discuss many other notable examples
of build systems, and their relationships,
in~\S\ref{sec-related}.

@@

\Make\footnote{There are numerous implementations of \Make and none comes with a
formal specification. In this paper we therefore use a simple and sensible
approximation to a real \Make that you might find on your machine.} was developed
more than 40 years ago to automatically build software libraries and executable
programs from source code. It uses \emph{makefiles} to describe tasks (often
referred to as \emph{build rules}) and their dependencies in a simple textual form.
For example\footnote{Do not worry about the dependency on \cmd{stdio.h}, and
its transitive (perhaps dynamic) dependencies.
We will deal with these in Section~\S\ref{sec-task-monad}.}:

@@

\noindent
The above makefile lists three tasks: (i) compile a utility library comprising
files \cmd{util.h} and \cmd{util.c} into \cmd{util.o} by
executing\footnote{In this example we pretend \cmd{gcc} is a pure function for the
sake of simplicity. In reality, there are multiple versions of \cmd{gcc} and the
actual binary that is used to compile and link files is often also listed as a task
dependency.} the command \cmd{gcc -c util.c}, (ii) compile the main source file
\cmd{main.c} into \cmd{main.o}, and (iii) link object files \cmd{util.o} and
\cmd{main.o} into the executable \cmd{main.exe}. The makefile contains the
complete information about the \emph{task dependency graph}, which is shown in
Fig.~\ref{fig-make}(a).

@@

To achieve minimality \Make relies on two main ideas: (i) it uses \emph{file
modification time} to detect which files changed\footnote{Technically, you
can fool \Make by altering the modification time of a file without changing its
content, e.g. by using the command \cmd{touch}. \Make is therefore minimal only
under the assumption that you do not do that.}, and (ii) it constructs a task
dependency graph from the information contained in the makefile and executes
tasks in a \emph{topological order}. For a more concrete description
see~\S\ref{sec-implementation-make}.

@@

\noindent
The formula in \cmd{B1} uses the \cmd{INDIRECT} function, which takes a string and
returns the value of the cell with that name.  The string evaluates
to \cmd{"A1"}, so \cmd{B1} evaluates to \cmd{10}. However, the dependencies of the
formula in \cmd{B1} are determined by the value of \cmd{C1}, so there is no
possibility\footnote{In this particular example one might say that
the value of \cmd{C1} is statically known, but imagine that it is the result of a long computation
chain -- its value will only become available during the build.}
of computing the dependency graph prior to calc.

To support dynamic dependencies,
\Excel's build algorithm~\cite{excel_recalc} is significantly different from
\Make. \Excel uses the \emph{calculation chain}
produced by the previous build as an approximation to the correct topological
order. During recalculation, \Excel processes cells in this order, but can
\emph{defer recalculation of a cell} by moving it down the chain if a newly
discovered dependency has not yet been rebuilt. We refer to this
algorithm as \emph{restarting}, and will discuss it in more detail
in~\S\ref{sec-implementation-excel}.

Dynamic dependencies complicate minimality.  In the above example, \cmd{B1}
should only be recomputed if \cmd{A1} or \cmd{C1} change, but not if (say)
\cmd{A2} changes; but these facts are not statically apparent. In practice
\Excel implements a conservative approximation to minimality: roughly, it
recomputes a formula if (i) the formula text mentions a changed cell, or (ii)
the formula uses a function like \cmd{INDIRECT} whose dependencies are not
statically visible.

Another distinguishing feature of \Excel is \emph{self-tracking}. Most build
systems only track changes of inputs and intermediate results, but \Excel can
also track changes in the tasks themselves: if a formula is modified, \Excel
will recompute it and propagate the changes. Self-tracking is uncommon in
software build systems, where one often needs to manually initiate a full
rebuild even if just a single build task has changed. We discuss
self-tracking further in~\S\ref{sec-tracking-aspects}.

@@

\begin{itemize}
    \item \cmd{LICENSE} is an input text file containing the project license.
    \item \cmd{release.txt} is a text file listing all files that should be in the release. This file
    is produced by concatenating input files \cmd{bins.txt} and \cmd{docs.txt}
    that list all binary and documentation files of the project.
    \item \cmd{release.tar} is the release archive built by executing the
    command \cmd{tar} on the release files.
\end{itemize}

@@

\Shake's implementation is different from both \Make and \Excel in two aspects.
First, it uses the dependency graph from the previous build to decide which
files need to be rebuilt. This idea has a long history, going back to
\emph{incremental}~\cite{demers1981incremental},
\emph{adaptive}~\cite{acar2002adaptive}, and
\emph{self-adjusting computations} (see \cite{acar2007selfadjusting} and \S\ref{sec-related}).
Second, instead of abandoning
and deferring the execution of tasks whose newly discovered dependencies have
not yet been built (as \Excel does), \Shake \emph{pauses} their execution until
the dependencies are brought up to date. We refer to this build algorithm as
\emph{recursive}.

>>>>

Build systems automate the execution of repeatable tasks for individual users
and large organisations. In this section we explore the design space of build
systems, using four concrete examples: \Make~\cite{feldman1979make},
\Shake~\cite{mitchell2012shake}, \Bazel~\cite{bazel}, and
\Excel~\cite{advanced_excel}\footnote{\Excel appears very different to the
others but, seen through the lens of this paper, it is very close indeed.}.
We have carefully chosen these four to illustrate the various axes on which
build systems differ; we discuss many other notable examples of build systems,
and their relationships, in~\S\ref{sec-related}.

@@

\Make\footnote{There are numerous implementations of \Make and none comes with a
formal specification. In this paper we therefore use a simple and sensible
approximation to a real \Make that you might find on your machine.} was developed
more than 40 years ago to automatically build software libraries and executable
programs from source code. It uses \emph{makefiles} to describe tasks (often
referred to as \emph{build rules}) and their dependencies in a simple textual form.
For example:

@@

\noindent
The above makefile lists three tasks: (i) compile a utility library comprising
files \cmd{util.h} and \cmd{util.c} into \cmd{util.o} by
executing\footnote{In this example we pretend \cmd{gcc} is a pure function for
the sake of simplicity. In reality, there are multiple versions of \cmd{gcc}
and the actual binary used to compile and link files, and standard libraries
such that \cmd{stdio.h}, are often also listed as dependencies.} the command
\cmd{gcc -c util.c}, (ii) compile the main source file \cmd{main.c} into
\cmd{main.o}, and (iii) link object files \cmd{util.o} and \cmd{main.o} into the
executable \cmd{main.exe}. The makefile contains the complete information about
the \emph{task dependency graph}, which is shown in Fig.~\ref{fig-make}(a).

@@

To achieve minimality \Make relies on two main ideas: (i) it uses \emph{file
modification times} to detect which files changed\footnote{Technically, you
can fool \Make by altering the modification time of a file without changing its
content, e.g. by using the command \cmd{touch}. \Make is therefore minimal only
under the assumption that you do not do that.}, and (ii) it constructs a task
dependency graph from the information contained in the makefile and executes
tasks in a \emph{topological order}. For a more concrete description
see~\S\ref{sec-implementation-make}.

@@

\noindent
The formula in \cmd{B1} uses the \cmd{INDIRECT} function, which takes a string
and returns the value of the cell with that name.  The string evaluates to
\cmd{"A1"}, so \cmd{B1} evaluates to \cmd{10}. However, the dependencies of the
formula in \cmd{B1} are determined by the value of \cmd{C1}, so there is no
possibility\footnote{In this particular example one might say that the value of
\cmd{C1} is statically known, but imagine that it is the result of a long
computation chain -- its value will only become available during the build.}
of computing the dependency graph before the build starts.

To support dynamic dependencies, \Excel's build algorithm~\cite{excel_recalc} is
significantly different from \Make. \Excel arranges the cells into a linear
sequence, called the \emph{calc chain}.  During the build, \Excel processes
cells in the calc-chain sequence, but if computing a cell \cmd{C} requires the
value of a cell \cmd{D} that has not yet been computed, \Excel \emph{aborts}
computation of \cmd{C}, moves \cmd{D} before \cmd{C} in the calc chain, and
resumes the build starting with \cmd{D}. When a build is complete, the resulting
calc chain respects all the dynamic dependencies of the spreadsheet. When an
input value (or formula) is changed, \Excel uses the final calc chain from the
\emph{previous} build as its starting point so that, in the common case where
changing an input value does not change dependencies, there are no aborts.
Notice that build always succeeds regardless of the initial calc chain (barring
truly circular dependencies); the calc chain is just an optimisation.
We refer to this algorithm as \emph{restarting}, and will discuss it
in more detail in~\S\ref{sec-implementation-excel}.

Dynamic dependencies complicate minimality.  In the above example, \cmd{B1}
should only be recomputed if \cmd{A1} or \cmd{C1} change, but not if (say)
\cmd{A2} changes; but these facts are not statically apparent. In practice
\Excel implements a conservative approximation to minimality: it recomputes a
formula if~(i)~the formula mentions a changed cell, or~(ii)~the formula uses a
function like \cmd{INDIRECT} whose dependencies are not statically visible,
or~(iii)~the formula itself has changed.

Item (iii) in the above list highlights a distinguishing feature of \Excel: it
is \emph{self-tracking}. Most build systems only track changes of inputs and
intermediate results, but \Excel also tracks changes in the tasks themselves: if
a formula is modified, \Excel will recompute it and propagate the changes.
Self-tracking is uncommon in software build systems, where one often needs to
manually initiate a full rebuild even if just a single build task has changed.
We discuss self-tracking further in~\S\ref{sec-tracking-aspects}.

@@

\begin{itemize}
    \item \cmd{LICENSE} is an input text file containing the project license.
    \item \cmd{release.txt} is a text file listing all files that should be in the release. This file
    is produced by concatenating input files \cmd{bins.txt} and \cmd{docs.txt},
    which list all binary and documentation files of the project.
    \item \cmd{release.tar} is the release archive built by executing the
    command \cmd{tar} on the release files.
\end{itemize}

@@

\Shake's implementation is different from both \Make and \Excel in two aspects.
First, it uses the dependency graph from the previous build to decide which
files need to be rebuilt. This idea has a long history, going back to
\emph{incremental}~\cite{demers1981incremental},
\emph{adaptive}~\cite{acar2002adaptive}, and
\emph{self-adjusting computations} (see \cite{acar2007selfadjusting} and
\S\ref{sec-related}). Second, instead of aborting and deferring the execution of
tasks whose newly discovered dependencies have not yet been built (as \Excel
does), \Shake \emph{suspends} their execution until the dependencies are brought
up to date. We refer to this build algorithm as \emph{suspending}.
